---
title: "Quiz 3"
author: "Alexander Tuzhikov"
date: "November 17, 2015"
output: 
  html_document: 
    fig_caption: yes
    fig_height: 9
    fig_width: 12
    highlight: haddock
    keep_md: yes
    number_sections: yes
    theme: cosmo
    toc: yes
---

#Q1
```{r question 1}
library(AppliedPredictiveModeling)
data(segmentationOriginal)
library(caret)
library(dplyr)

#Subset the data to a training set and testing set based on the Case variable in the data set. 
training<- segmentationOriginal %>% filter(Case == "Train")
testing<- segmentationOriginal %>% filter(Case=="Test")

#Set the seed to 125 and fit a CART model with the rpart method using all predictor variables and default caret settings. 
set.seed(125)
model<- train(Class ~ ., method="rpart", data=training)

#In the final model what would be the final model prediction for cases with the following variable values:

model$finalModel

library(rattle)
fancyRpartPlot(model$finalModel)
```

#Q2

If K is small in a K-fold cross validation is the bias in the estimate of out-of-sample (test set) accuracy smaller or bigger? If K is small is the variance in the estimate of out-of-sample (test set) accuracy smaller or bigger. Is K large or small in leave one out cross validation?

* The bias is larger and the variance is smaller. Under leave one out cross validation K is equal to the sample size.

#Q3
```{r question 3}
library(pgmm)
data(olive)
olive = olive[,-1]

model<- train(Area ~., data=olive, method="rpart")
newdata = as.data.frame(t(colMeans(olive)))
predict(model, newdata)
```

#Q4
```{r question 4}
library(ElemStatLearn)
data(SAheart)
set.seed(8484)
train = sample(1:dim(SAheart)[1],size=dim(SAheart)[1]/2,replace=F)
trainSA = SAheart[train,]
testSA = SAheart[-train,]

set.seed(13234)
logitModel <- train(chd ~ age + alcohol + obesity + tobacco + 
                      typea + ldl, data=trainSA, method="glm", 
                    family="binomial")
logitModel
missClass <- function(values,prediction){sum(((prediction > 0.5)*1) != values)/length(values)}
predictTrain <- predict(logitModel, trainSA)
predictTest <- predict(logitModel, testSA)
missClass(trainSA$chd, predictTrain) 
missClass(testSA$chd, predictTest)
```

#Q5
```{r}
library(ElemStatLearn)
library(randomForest)
data(vowel.train)
data(vowel.test)
vowel.train$y <- as.factor(vowel.train$y)
vowel.test$y <- as.factor(vowel.test$y)
set.seed(33833)
modelRf <- randomForest(y ~ ., data = vowel.train, importance = FALSE)
order(varImp(modelRf), decreasing=T)
```